Hilbert space methods to approximate Gaussian processes using Stan

### 2019-02-26 Paper/manuscript/manuscript_05.tex
review Paul changes

Sparse GPs are based on building low-rank approximations of the covariance matrix of the complete data by means of 'inducing variables' (Paul: if you introduce this term, you may need to explain it. Alternatively, can we just get rid of it?). They reduce the dimension of the posterior distribution thereby reducing the memory requirements to $O(nm)$ and computational complexity to $O(nm^2)$ for some $m < n$. This approach still requires Cholesky factorization although now based on a smaller covariance matrix. While computational complexity is reduced considerably, ensuring convergence to the exact GP model is not necessarily straightforward (Paul: this may need a citation). As such, reducing the amount of data while maintaining the correct posterior distribution is not a simple procedure. A unifying view on sparse GPs based on approximate generative methods is provided in \cite{quinonero2005unifying}, while a general review can be found in \cite{rasmussen2006gaussian}. More recent developments in the context of sparse GPs include points-based sparse approximation methods \citep{wilson2015kernel}, which have been further developed by Bui et. al (2017) (Paul: cite this study correctly in latex) to perform approximations at inference time rather than at modeling time.

@@

Yet another approach to approximate GPs -- on which we will focus in the present paper -- was proposed by \cite{solin2018hilbert} who suggested to use reduced-rank approximations of the covariance kernel. This method decomposes the kernel into basis functions which represent the GP as a linear model and thus makes its estimation considerably faster. In principal, we are free in the choice of basis functions, but the Laplace eigenfunctions were found to be particularily appealing \citep{solin2018hilbert}. Not only can they be computed analytically, but they are also independent of the particular choice of the covariance kernel including the hyperparameters. The latter comes with several advantages when performing inference via algorithms that require differentiation of the posterior density (Paul: cite something here). In a fully Bayesian framework based on sampling methods, the proposed approximate GP model has a computational complexity of just $O(nm+m)$, with $m<<n$ in every sampling step. The posterior distribution of the reduced-rank GP approximation is of dimension $m$ and thus considerably smaller than in the case of exact GPs. This not only lowers the memory requirements but also reduces the correlations between latent values and the corresponding hyperparameters of the GP, which can help to further improve sampling efficiency.

While \cite{solin2018hilbert} fully developed the mathematical theory behind reduced-rank approximations of GPs, they did not put much effort in analyzing the performance and accuracy of the method in relation to key factors such as the number of basis functions, desired prediction space (Paul: I replace 'box size' because, at that point, readers will likely have no idea what that means), or properties of the true functional relationship between covariates and response variable. However, for practical implementations of approximate GPs, the latter factors are critical and thus require further investigation. In the present paper, we will fill this gap by providing practical recommendations for the choice of these factors based on the recognized relations among them. What is more, we provide intuitive visualizations of these relations that will help users to improve performance and save computation time, while at the same time maintaining close approximations of exact GPs.

Although there are several GP specific software packages available to date (Paul: cite those), each provide efficient implementations only for a restricted range of GP based models. In this paper we do not focus on the fastest possible inference for a specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks such as Stan (cite). To this end, we want to make full Bayesian inference over Gaussian process models accessible to a broader audience by making them easy, fast and save to apply.

@@

(Paul: The content and the purpose of the following sentences is not fully clear to me. Do we need them in the Introduction or can they be removed, or put into the next section?) While Sparse Spectrum GP is based on a sparse spectrum, the reduced-rank method proposed in this paper aims to make the spectrum as ‘full’ as possible at a given rank. Recent Splines models can reproduce the Matern family of covariance functions (see, e.g., \cite{wood2003thin}), however our approach can reproduce basically all of the stationary covariance functions.

@@

A Gaussian process (GP) is a stochastic process which defines the distribution over a collection of random variables indexed by a continuous variable, i.e. $\left\lbrace f(t): t \in \mathcal{T}\right\rbrace$ for some index set $\mathcal{T}$. Gaussian processes have the defining property that the marginal distribution of every finite subset of random variables, $\left\lbrace f(t_1), f(t_2), \hdots, f(t_K) \right\rbrace$, is a multivariate Gaussian distribution.
\\
\\
In this work, Gaussian processes will take the role of prior distributions over function spaces for non-parametric regression in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \mathbf{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is a noisy observation of an unknown function $f$ at $\mathbf{x}_n \in \mathbb{R}^D$ , i.e. $y_n = f(\mathbf{x}_n) + \epsilon_n$. Our goal is to predict the value of the function $f^*=f(\mathbf{x}^*)$ evaluated at a new input point $\mathbf{x}^* \in \mathbb{R}^D$. That is, we want to obtain the predictive distribution $p(f^*|\mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.
\\
\\
We assume a Gaussian process prior for the collection of function values $\mathbf{f}=\left\lbrace f(\mathbf{x}_n) \right\rbrace_{n=1}^N $, $\mathbf{f} \sim \mathcal{GP}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x}'))$, where $\mu: \mathbb{R}^D \rightarrow \mathbb{R}$ and $k: \mathbb{R}^D \times \mathbb{R}^D \rightarrow \mathbb{R}$ are the mean and covariance functions, respectively,
%
\begin{align*}
 	\mu(\mathbf{x}) &= \mathbb{E}\!\left[f(\mathbf{x})\right],\\ 
 	k(\mathbf{x}, \mathbf{x}') &= \mathbb{E}\!\left[\left( f(\mathbf{x}) - \mu(\mathbf{x}) \right)\left( f(\mathbf{x}') - \mu(\mathbf{x}') \right)\right].
\end{align*}

The mean and covariance functions represent how the random functions behave on average and how the different points in the input space co-vary with respect to each other. The mean and covariance functions completely characterize the multivariate Gaussian for $\mathbf{f}$, 
\begin{align*}
\mathbf{f} \sim \mathcal{N}(\boldsymbol{\mu},K),
\end{align*}

\noindent where $\boldsymbol{\mu} = \left\lbrace \mu(\mathbf{x}_n) \right\rbrace_{n=1}^N$ is the multivariate mean and $K$ the covariance matrix, where the element $K_{i,j}$ is given by the covariance function $k(\mathbf{x}_i,\mathbf{x}_j)$ at input values $\mathbf{x}_i$ and $\mathbf{x}_j$. The joint distribution of $\mathbf{f}$ and $f^*$ is also a multivariate Gaussian as,
%
\begin{eqnarray*}
p(\mathbf{f},\mathbf{f}')=\mathcal{N} \left( \left[ \begin{array}{cc}
\mathbf{f} \\ 
\mathbf{f}'
\end{array} \right] | 0,\left[ \begin{array}{cc}
K_{\mathbf{f},\mathbf{f}} & K_{\mathbf{f},f^*} \\ 
K_{f^*,\mathbf{f}} & K_{f^*,f^*}
\end{array} \right] \right),
\end{eqnarray*}

@@

\vspace{0.2cm}
In a $D$-dimensional input space the total number of eigenfunctions and eigenvalues in the approximation corresponds to the total number of $D$-tuples over $J$, which is $\prod_{d=1}^{D} J_d$, where $J_d$ is the number of basis function for the dimension $d$. Let $\mathbb{S}\in {\rm I\!R}^{\prod_{d=1}^{D} J_d \times D}$ be the set of $D$-tuples elements. Each new eigenfunction $\phi^{\ast}_j$ correspond to the product of the univariate eigenfunctions whose indices corresponds to the the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$. And each new eigenvalue $\lambda^{\ast}_j$ is a $D$-vector whose elements are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$. Let $j=\{1,\cdots,J^D\}$,
%
\begin{eqnarray} \label{bf&lambda_multi}
\begin{split}
\phi^{\ast}_j(\mathbf{x}) &=& \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(\mathbf{x}_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \text{sin}\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(\mathbf{x}_d+L_d)\right) \\
%
\lambda^{\ast}_j &=& \{\lambda_{\mathbb{S}_{jd}} \}_{d=1}^D =  \{ \left(\tfrac{\pi \mathbb{S}_{jd}}{2L_d}\right)^2 \}_{d=1}^D  
\end{split}
\end{eqnarray}

>>>>

Sparse GPs are based on building low-rank approximations of the covariance matrix of the complete data by means of selecting a subset $m$ of training points on which to base computation. They reduce the dimension of the posterior distribution thereby reducing the memory requirements to $O(nm)$ and computational complexity to $O(nm^2)$ for some $m < n$. This approach still requires Cholesky factorization although now based on a smaller covariance matrix. While computational complexity is reduced considerably, ensuring convergence to the exact GP model is not necessarily straightforward (Paul: this may need a citation). As such, reducing the amount of data while maintaining the correct posterior distribution is not a simple procedure. A unifying view on sparse GPs based on approximate generative methods is provided in \cite{quinonero2005unifying}, while a general review can be found in \cite{rasmussen2006gaussian}. (Gabi: Could you choose between one of these two set of sentences in different colors?) \textcolor{blue}{More recent developments in the context of sparse GPs include points-based sparse approximation methods \citep{wilson2015kernel}, which have been further developed by \citep{bui2017unifying} to perform approximations at inference time rather than at modeling time.} \textcolor{red} {New developments in the context of sparse GPs include a structured kernel interpolation (SKI) method \citep{wilson2015kernel}. More recently \citep{bui2017unifying} revisit the inducing points-based sparse approximation methods to perform approximations at inference time rather than at modeling time.}

@@

Yet another approach to approximate GPs -- on which we will focus in the present paper -- was proposed by \cite{solin2018hilbert} who suggested to use reduced-rank approximations of the covariance kernel. This method decomposes the kernel into basis functions which represent the GP as a linear model and thus makes its estimation considerably faster. In principal, we are free in the choice of basis functions, but the Laplace eigenfunctions were found to be particularily appealing \citep{solin2018hilbert}. Not only can they be computed analytically, but they are also independent of the particular choice of the covariance kernel including the hyperparameters. The latter comes with several advantages when performing inference via algorithms that require differentiation of the posterior density (Paul: cite something here). In a fully Bayesian framework based on sampling methods, the proposed approximate GP model has a computational complexity of just $O(nm+m)$, with $m<<n$ in every sampling step. The posterior distribution of the reduced-rank GP approximation is of dimension $m$ and thus considerably smaller than in the case of exact GPs. This not only lowers the memory requirements and computational complexity but also reduces the correlations between latent values and the corresponding hyperparameters of the GP, which can help to further improve sampling efficiency.

While \cite{solin2018hilbert} fully developed the mathematical theory behind reduced-rank approximations of GPs, they did not put much effort in analyzing the performance and accuracy of the method in relation to key factors such as the number of basis functions, desired prediction space, or properties of the true functional relationship between covariates and response variable. However, for practical implementations of approximate GPs, the latter factors are critical and thus require further investigation. In the present paper, we will fill this gap by providing practical recommendations for the choice of these factors based on the recognized relations among them. What is more, we provide intuitive visualizations of these relations that will help users to improve performance and save computation time, while at the same time maintaining close approximations of exact GPs.

Although there are several GP specific software packages available to date (Paul: cite those. Gabi: This full paragraph comes from Aki. I think he refers to some Splines methods and maybe others approaches as well), each provide efficient implementations only for a restricted range of GP based models. In this paper we do not focus on the fastest possible inference for a specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks such as Stan \citep{carpenter2017stan}. To this end, we want to make full Bayesian inference over Gaussian process models accessible to a broader audience by making them easy, fast and save to apply.

@@

(Paul: The content and the purpose of the following sentences is not fully clear to me. Do we need them in the Introduction or can they be removed, or put into the next section? Gabi: In this paragraph I tried to state two differences I found in the literature between our method and the Sparse Spectrum GP and the Splines method. But I not know where locate them.) While Sparse Spectrum GP is based on a sparse spectrum, the reduced-rank method proposed in this paper aims to make the spectrum as ‘full’ as possible at a given rank. Recent Splines models can reproduce the Matern family of covariance functions (see, e.g., \cite{wood2003thin}), however our approach can reproduce basically all of the stationary covariance functions.

@@

A Gaussian process (GP) is a stochastic process which defines the distribution over a collection of random variables indexed by a continuous variable, i.e. $\left\lbrace f(t): t \in \mathcal{T}\right\rbrace$ for some index set $\mathcal{T}$. Gaussian processes have the defining property that the marginal distribution of every finite subset of random variables, $\left\lbrace f(t_1), f(t_2), \hdots, f(t_K) \right\rbrace$, is a multivariate Gaussian distribution.
\\
\\
In this work, Gaussian processes will take the role of prior distributions over function spaces for non-parametric regression in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \mathbf{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is a noisy observation of an unknown function $f$ at $\mathbf{x}_n \in \mathbb{R}^D$ , i.e. $y_n = f(\mathbf{x}_n) + \epsilon_n$. Our goal is to predict the value of the function $f^*=f(\mathbf{x}^*)$ evaluated at a new input point $\mathbf{x}^* \in \mathbb{R}^D$. That is, we want to obtain the predictive distribution $p(f^*|\mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.
\\
\\
We assume a Gaussian process prior for $f \sim \mathcal{GP}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x}'))$, where $\mu: \mathbb{R}^D \rightarrow \mathbb{R}$ and $k: \mathbb{R}^D \times \mathbb{R}^D \rightarrow \mathbb{R}$ are the mean and covariance functions, respectively,
%
\begin{align*}
 	\mu(\mathbf{x}) &= \mathbb{E}\!\left[f(\mathbf{x})\right],\\ 
 	k(\mathbf{x}, \mathbf{x}') &= \mathbb{E}\!\left[\left( f(\mathbf{x}) - \mu(\mathbf{x}) \right)\left( f(\mathbf{x}') - \mu(\mathbf{x}') \right)\right].
\end{align*}

The mean and covariance functions completely characterize the Gaussian process prior, and represent how the random functions behave on average and how the different points in the input space co-vary with respect to each other, respectively. The Gaussian process prior can be interpreted such as any finite number of function values $\mathbf{f}=\left\lbrace f(\mathbf{x}_n) \right\rbrace_{n=1}^N$ is multivariate Gaussian, 
\begin{align*}
\mathbf{f} \sim \mathcal{N}(\boldsymbol{\mu},K),
\end{align*}

\noindent where $\boldsymbol{\mu} = \left\lbrace \mu(\mathbf{x}_n) \right\rbrace_{n=1}^N$ is the multivariate mean and $K$ the covariance matrix, where the element $K_{i,j}$ are formed by evalating the covariance function $k(\mathbf{x}_i,\mathbf{x}_j)$ at input values $\mathbf{x}_i$ and $\mathbf{x}_j$. The joint distribution of $\mathbf{f}$ and $f^*$ is also a multivariate Gaussian as,
%
\begin{eqnarray*}
p(\mathbf{f},\mathbf{f}')=\mathcal{N} \left( \left[ \begin{array}{cc}
\mathbf{f} \\ 
\mathbf{f}'
\end{array} \right] | 0,\left[ \begin{array}{cc}
K_{\mathbf{f},\mathbf{f}} & K_{\mathbf{f},f^*} \\ 
K_{f^*,\mathbf{f}} & K_{f^*,f^*}
\end{array} \right] \right),
\end{eqnarray*}

@@

\vspace{0.2cm}
In a $D$-dimensional input space the total number of eigenfunctions and eigenvalues in the approximation corresponds to the total number of $D$-tuples over $J$, which is $\prod_{d=1}^{D} J_d$, where $J_d$ is the number of basis function for the dimension $d$. Let $\mathbb{S}\in {\rm I\!R}^{\prod_{d=1}^{D} J_d \times D}$ be the set of $D$-tuples elements. Each new eigenfunction $\phi^{\ast}_j$ correspond to the product of the univariate eigenfunctions whose indices corresponds to the the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$. And each new eigenvalue $\lambda^{\ast}_j$ is a $D$-vector whose elements are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$. Let $j=\{1,\cdots,J^D\}$,
%
\begin{align} \label{bf&lambda_multi}
\begin{split}
\phi^{\ast}_j(\mathbf{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(\mathbf{x}_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \text{sin}\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(\mathbf{x}_d+L_d)\right) \\
%
\lambda^{\ast}_j &= \{\lambda_{\mathbb{S}_{jd}} \}_{d=1}^D =  \{ \left(\tfrac{\pi \mathbb{S}_{jd}}{2L_d}\right)^2 \}_{d=1}^D  
\end{split}
\end{align}
